---
title: "R_finalproject"
output:
  pdf_document: default
  word_document: default
date: "2026-04-27"
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
# This sets the working directory for ALL chunks in this file
knitr::opts_knit$set(root.dir = "/usr4/ds596/vgupta7/final_project")
```

Note to self:
When starting a new R studio SCC session, run the library load chunk, skip GDCquery, GDCdownload, and DESeq() chunks, and load("brca_deseq_results.rda")

# Install and Load Bioconductor 
```{r}
#if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
#BiocManager::install(c("TCGAbiolinks", "DESeq2", "SummarizedExperiment"))
```

# Setup and Loading libraries 
```{r}
#Data manipulation
library(tidyverse)

#Bioinformatics specific packages
library(TCGAbiolinks)
library(DESeq2)
library(SummarizedExperiment)

# Setting the working directory
setwd("/usr4/ds596/vgupta7/final_project")
``` 

#PART 1: DATA ACQUISITION

# Search for data on the GDC server
```{r}
# Define the search for BRCA Tumor vs Normal samples
query_brca <- GDCquery(
    project = "TCGA-BRCA",
    data.category = "Transcriptome Profiling",
    data.type = "Gene Expression Quantification",
    workflow.type = "STAR - Counts",
    sample.type = c("Primary Tumor", "Solid Tissue Normal")
)

# View the results of your search
output_query <- getResults(query_brca)

# Check how many of each sample type you found
table(output_query$sample_type)
```
The TCGA-BRCA cohort has 1,111 tumor samples and 113 normal tissue samples.


```{r}
# Download the actual data files to your SCC directory
GDCdownload(query_brca)
```

# Prepare the GDC data
```{r}
#Prepare the data into a SummarizedExperiment object
data_brca <- GDCprepare(query_brca)
```

R is reading every single one of the 1,224 files and extracting the gene names and "STAR - Counts" column from each file. Then it's aligning them all into one massive matrix where every row is a gene and every column is one patient. Finally, it attaches the clinical metadata (survival info, "Tumor vs. Normal" labels, etc) to that matrix.

## Save the object
```{r}
save(data_brca, file = "brca_prepared_data.rda")
```

## Have a look at the data
```{r}
# See the dimensions (Rows = Genes, Cols = Samples)
dim(data_brca)

# See the first few "Normal" vs "Tumor" labels
table(colData(data_brca)$sample_type)
```
There are 60,660 genes and 1,224 samples (this is the full transcriptomic profile for the BRCA cohort).

#PART 2: DATA MODELING

```{r}
# 1. Convert the data to a DESeq object
# Use 'sample_type' as the variable to compare
dds <- DESeqDataSet(data_brca, design = ~ sample_type)

# 2. Pre-filtering
# This removes genes with almost zero reads, which speeds up the math and makes the results more reliable.
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]

# 3. RUN THE ANALYSIS
# This performs normalization, dispersion estimation, and the actual testing
dds <- DESeq(dds)

# 4. Get the results
# This creates a table with Log2FoldChange and p-values
res <- results(dds, contrast = c("sample_type", "Primary Tumor", "Solid Tissue Normal"))

# 5. Save the results so you don't have to run this again
save(res, dds, file = "brca_deseq_results.rda")

```
This chunk of code calculates the p-values and fold changes that will be used for the Volcano Plot.

Note: I hit my 10 GB quota while this ran so the deseq results didn't properly save.

# Check the results of the differential expression analysis
```{r}
summary(res)
```

# Save results into a data frame 
```{r}
# Convert the results to a simple, small data frame
res_df <- as.data.frame(res)

# Save only the results table
save(res_df, file = "brca_results_dataframe.rda")

# Save the full DESeq objects again (the "Heavy" model)
# Doing this again because I hit an error with this earlier (exceeded quota)
save(dds, res, file = "brca_deseq_results_final.rda")
```

# Verify that the save worked
```{r}
# Try to load the file back in under a different name to test it
load("brca_results_dataframe.rda")
head(res_df)
```
Success!

# Identify the top 10 most up-regulated genes
```{r}
# Filter for significant genes and sort by Fold Change
top_hits <- res_df %>%
  filter(padj < 0.05) %>%
  arrange(desc(log2FoldChange)) %>%
  head(10)

print(top_hits)

```
Those are high log2FoldChange values! This means that those genes (listed as Ensembl IDs) are massively up-regulated in tumor cells compared to normal tissue.

# PART 3: DATA VISUALIZATION 

# Figure 1: Volcano Plot

# Install EnhancedVolcano
```{r}
#if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
#BiocManager::install("EnhancedVolcano")

library(EnhancedVolcano)

```

# Map Ensembl IDs to Gene Symbols
```{r}
# Install the mapping database
#BiocManager::install("org.Hs.eg.db")
#library(org.Hs.eg.db)

# 1. Strip the version numbers (the ".13" at the end of IDs)
# DESeq2 sometimes adds these, but the database needs just the ENSG ID
clean_ids <- gsub("\\..*", "", rownames(res_df))

# 2. Map IDs to Symbols
mapping <- mapIds(org.Hs.eg.db,
                  keys = clean_ids,
                  column = "SYMBOL",
                  keytype = "ENSEMBL",
                  multiVals = "first")

# 3. Add the symbols to your results dataframe
res_df$symbol <- mapping
```
This failed because I'm approaching the quota again.

# Manually label the top 10 up-regulated genes
```{r}
# 1. Create a column that is empty by default
res_df$presentation_label <- ""

# 2. Add the top hits 
res_df["ENSG00000198930.13", "presentation_label"] <- "AZGP1"
res_df["ENSG00000135346.9", "presentation_label"] <- "CASP14"
res_df["ENSG00000197172.10", "presentation_label"] <- "MTRNR2L2"
res_df["ENSG00000101441.4", "presentation_label"] <- "PIP"
res_df["ENSG00000213401.10", "presentation_label"] <- "AL022318.1"
res_df["ENSG00000198788.9", "presentation_label"] <- "MTRNR2L1"
res_df["ENSG00000259527.2", "presentation_label"] <- "AC011477.1"
res_df["ENSG00000221867.9", "presentation_label"] <- "AL133341.1"
res_df["ENSG00000198681.7", "presentation_label"] <- "MTRNR2L8"
res_df["ENSG00000123500.10", "presentation_label"] <- "MUC1"

```

# Identify the top down-regulated genes
```{r}
# Sort by the most NEGATIVE fold change
down_regulated <- res_df %>%
  filter(padj < 0.05) %>%
  arrange(log2FoldChange) %>%
  head(5)

print(down_regulated)
```
```{r}

# Find the 5 most down-regulated genes in YOUR results
actual_down <- res_df %>%
  filter(padj < 0.01 & log2FoldChange < -2) %>%
  arrange(log2FoldChange) %>%
  head(5)

# Print them to your console so you can see the names
print(actual_down)

# Update the labels using the IDs found above
# Replace 'ID_1', 'ID_2', etc., with the IDs printed from the step above
#res_df[rownames(actual_down)[1], "presentation_label"] <- "Down_1" 
# (And so on for the others)
```

# Manually label the top 5 down-regulated genes
```{r}
res_df["ENSG00000034971.17", "presentation_label"] <- "FXYD1"
res_df["ENSG00000174697.5", "presentation_label"] <- "LEP"    
res_df["ENSG00000149124.11", "presentation_label"] <- "CASP5"
res_df["ENSG00000277737.3", "presentation_label"]  <- "LINC02224"
res_df["ENSG00000287148.1", "presentation_label"]  <- "AC009283.1"

```

# Create the Volcano Plot
```{r}
# Open the image file
#png("BRCA_Final_Volcano.png", width = 2400, height = 2000, res = 300)

#Combine the top up-regulated and down-regulated hits
selected_labels <- c("AZGP1", "CASP14", "PIP", "MUC1", "MTRNR2L1", 
                        "FXYD1", "LEP", "CASP5", "LINC02224")

EnhancedVolcano(res_df,
    lab = res_df$presentation_label,
    x = 'log2FoldChange',
    y = 'padj',
    selectLab = selected_labels,
    title = 'TCGA-BRCA Transcriptomic Profile',
    subtitle = 'Primary Tumor vs. Solid Tissue Normal',
    titleLabSize = 15,
    subtitleLabSize = 11,
    pCutoff = 0.01,
    FCcutoff = 2.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2'),
    legendLabels = c('Not Significant', 'High Fold Change (NS)', 
                   'Significant (Small FC)', 'Top Hits (Sig. & High FC)'),
    legendPosition = 'right',
    legendLabSize = 10,
    drawConnectors = TRUE,
    widthConnectors = 0.5,
    max.overlaps = 20)

```


```{r}
save(res_df, file = "brca_final_table_for_plots.rda")

```

Starting a new session, load deseq results to generate the rest of the figures
```{r}
# Load the processed results
load("brca_results_dataframe.rda")

# Load the full model (for the Heatmap and Boxplots)
load("brca_deseq_results_final.rda")

```

# Set working directory
```{r}
setwd("/usr4/ds596/vgupta7/final_project")

```

# Figure 2: PCA plot

# Transform the data for PCA
```{r}
# Apply VST transformation
# blind = FALSE means it will use your design formula to be more accurate
vsd <- vst(dds, blind = FALSE)
```

Raw counts in RNA-seq have a "mean-variance" relationship (highly expressed genes have higher variance), which can bias the PCA. The Variance Stabilizing Transformation (VST) makes this look more like a normal distribution.

# Generate the PCA Plot
```{r}
# Generate the PCA data
pcaData <- plotPCA(vsd, intgroup = "definition", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

# Create the plot
library(ggplot2)
ggplot(pcaData, aes(PC1, PC2, color = definition)) +
  geom_point(size = 3, alpha = 0.6) + # Alpha makes the 1,200 points readable
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  theme_bw() + 
  scale_color_manual(values = c("Primary solid Tumor" = "red2", 
                                "Solid Tissue Normal" = "royalblue")) +
  labs(title = "Global Transcriptomic Variance (PCA)",
       subtitle = "TCGA-BRCA: Tumor vs. Normal",
       color = "Sample Type") +
  theme(plot.title = element_text(size = 15, face = "bold"),
        legend.position = "bottom")
```
# Figure 3: Heatmap

# Prepare data for heatmap (extract normalized counts for top genes)
```{r}
#if (!require("BiocManager", quietly = TRUE))
#install.packages("BiocManager")

#BiocManager::install("ComplexHeatmap")

library(ComplexHeatmap)
library(circlize) # For color mapping

# 1. Get the top 50 genes by adjusted p-value
top_genes <- res_df %>%
  arrange(padj) %>%
  head(50) %>%
  rownames()

# 2. Extract the VST-normalized counts for these genes
mat <- assay(vsd)[top_genes, ]

# 3. Z-score the rows (Standardize)
# This centers the data so '0' is the average expression for that gene.
# Red = Above average, Blue = Below average.
mat <- t(scale(t(mat))) 

# 4. Create an annotation bar for the top (Tumor vs Normal)
df <- as.data.frame(colData(dds)[, c("definition")])
colnames(df) <- "Type"
ha <- HeatmapAnnotation(df = df, 
                        col = list(Type = c("Primary solid Tumor" = "red2", 
                                            "Solid Tissue Normal" = "royalblue")))

```

# Generate the heatmap
```{r}
# Define the color palette (Blue to White to Red)
col_fun = colorRamp2(c(-2, 0, 2), c("royalblue", "white", "red2"))

# Create the Heatmap
heatmap_plot <- Heatmap(mat, 
        name = "Z-score", 
        top_annotation = ha,
        cluster_rows = TRUE, 
        cluster_columns = TRUE, # This groups similar samples together
        show_column_names = FALSE,
        show_row_names = TRUE, # Shows your Top 50 genes
        row_names_gp = gpar(fontsize = 8),
        column_title = "Top 50 Differentially Expressed Genes",
        col = col_fun)

# Draw it
draw(heatmap_plot)

```
This heatmap looks a bit cluttered, mostly because there is a large number of samples. So I'm going to make another one that only uses 200 samples to better see the pattern.


# Install ID mapping database 

```{r}
#BiocManager::install("org.Hs.eg.db")
library(org.Hs.eg.db)

```

# 1. Prepare Data and Map IDs
```{r}

# Clean the rownames for BOTH objects to remove version dots (e.g., .13, .5)
rownames(vsd) <- gsub("\\..*", "", rownames(vsd))
res_df$clean_id <- gsub("\\..*", "", rownames(res_df))

# Map IDs to Symbols for the vsd object
symbols <- mapIds(org.Hs.eg.db,
                  keys = rownames(vsd),
                  column = "SYMBOL",
                  keytype = "ENSEMBL",
                  multiVals = "first")

# Create the clean labels
clean_labels <- ifelse(is.na(symbols), rownames(vsd), symbols)
rownames(vsd) <- clean_labels
res_df$presentation_label <- clean_labels

# Get top 50 genes (now with symbols)
top_genes <- res_df %>%
  arrange(padj) %>%
  head(50) %>%
  pull(presentation_label)

# Extract main data matrix for these 50 genes
mat <- assay(vsd)[top_genes, ]
mat <- t(scale(t(mat))) # Z-score Standardization

```


# 2. Downsampling (100 Healthy vs. 100 Tumor)
```{r}

# Get full metadata from the dds object
coldata_df <- as.data.frame(colData(dds))

# Separate the sample groups
tumor_samples <- rownames(coldata_df[coldata_df$definition == "Primary solid Tumor", ])
normal_samples <- rownames(coldata_df[coldata_df$definition == "Solid Tissue Normal", ])

# Randomly select 100 samples from each, ensuring we don't pick more than exist
# Use a seed for reproducibility
set.seed(42)
select_tumor <- sample(tumor_samples, 100)
select_normal <- sample(normal_samples, 100)

# Combine selected samples
selected_cols <- c(select_tumor, select_normal)

# Subset both the matrix and metadata for the final plot
mat_sub <- mat[, selected_cols]

#Create the annotation for just these 200 samples
df_sub <- data.frame(Type = coldata_df[selected_cols, "definition"])
rownames(df_sub) <- selected_cols

```


# 3. Define annotations and generate new heatmap
```{r}

# Define the colors for the annotation bar
ha <- HeatmapAnnotation(df = df_sub,
                        col = list(Type = c("Primary solid Tumor" = "red2",
                                            "Solid Tissue Normal" = "royalblue")),
                        show_legend = TRUE,
                        annotation_name_gp = gpar(fontsize = 10, face = "bold"))

# Setup color mapping for the Z-score (Blue-White-Red)
col_fun = colorRamp2(c(-2, 0, 2), c("royalblue", "white", "red2"))

# Create the Heat-map
heatmap_final <- Heatmap(mat_sub,
        name = "Z-score", 
        top_annotation = ha,
        cluster_rows = TRUE, 
        cluster_columns = TRUE,
        show_column_names = FALSE,
        column_km = 2,
        row_names_gp = gpar(fontsize = 5),
        column_title = "Molecular Signature: Top 50 BRCA Genes (n=200)",
        col = col_fun)

# Draw the final plot
draw(heatmap_final, merge_legend = TRUE)

```

# Figure 4: Boxplots 

# First, let's look at the clinical metadata
```{r}
# View the first few rows of metadata
col_metadata <- as.data.frame(colData(dds))
colnames(col_metadata)

```
```{r}
table(col_metadata$paper_BRCA_Subtype_PAM50, useNA = "always")
```

```{r}
# Specifically look for common TCGA clinical variables
clinical_vars <- c("ajcc_pathologic_stage", "vital_status", "age_at_index", "paper_BRCA_Subtype_PAM50")
col_metadata[, colnames(col_metadata) %in% clinical_vars] %>% head()

```

# Figure 4: Boxplots

# Re-run the Transformation (forgot to save it last time!)

```{r}
# 1. Re-create the Variance Stabilized object
# This is necessary because 'assay()' needs the transformed matrix
vsd <- vst(dds, blind = FALSE)

# 2. Clean the IDs to Gene Symbols (just like we did for the heatmap)
# This ensures you can call genes by name (like "MMP11") instead of ENSG codes
library(org.Hs.eg.db)
rownames(vsd) <- gsub("\\..*", "", rownames(vsd))
symbols <- mapIds(org.Hs.eg.db,
                  keys = rownames(vsd),
                  column = "SYMBOL",
                  keytype = "ENSEMBL",
                  multiVals = "first")

# Update vsd with clean names
rownames(vsd) <- ifelse(is.na(symbols), rownames(vsd), symbols)

```
# Save the vsd object (in case I need to use it again)
```{r}
save(vsd, file = "brca_vst_transformed.rda")

```

# Generate the boxplot 

```{r}
library(ggplot2)

# 1. Prepare the data
# We'll pull the counts for MMP11 and combine with PAM50 labels
gene_to_plot <- "MMP11"
boxplot_data <- data.frame(
  Expression = assay(vsd)[gene_to_plot, ],
  Subtype = colData(dds)$paper_BRCA_Subtype_PAM50
)

# 2. Filter out NAs and the 'Normal' subtype (since we already compared to Normal)
boxplot_data <- boxplot_data %>%
  filter(!is.na(Subtype), Subtype != "Normal")

# 3. Create the plot
ggplot(boxplot_data, aes(x = Subtype, y = Expression, fill = Subtype)) +
  geom_boxplot(outlier.shape = NA) + # Hide outliers to keep it clean
  geom_jitter(width = 0.2, alpha = 0.2, size = 0.5) + # Show the 1,000+ points subtly
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  labs(title = paste("Expression of", gene_to_plot, "by PAM50 Subtype"),
       subtitle = "TCGA-BRCA Clinical Deep-Dive",
       x = "Molecular Subtype",
       y = "VST-Normalized Expression") +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 14))

```

# Remake the heatmap with Viridis colors for gene expression (to make it easier to understand)

```{r}
#Load the vst results
load("brca_vst_transformed.rda")
```

# Recreate the subsampled matrix
```{r}

# Clean the results IDs to match the vsd names (Symbols)
res_df$clean_id <- gsub("\\..*", "", rownames(res_df))

# Map symbols to res_df if you haven't already
library(org.Hs.eg.db)

res_symbols <- mapIds(org.Hs.eg.db, keys = res_df$clean_id, column = "SYMBOL", keytype = "ENSEMBL", multiVals = "first")
res_df$presentation_label <- ifelse(is.na(res_symbols), res_df$clean_id, res_symbols)

# Get top 50 genes
top_genes <- res_df %>% arrange(padj) %>% head(50) %>% pull(presentation_label)

# Extract and Scale
mat <- assay(vsd)[top_genes, ]
mat <- t(scale(t(mat)))

# Down sample (using seed 42)
set.seed(42)
coldata_df <- as.data.frame(colData(dds))
tumor_samples <- rownames(coldata_df[coldata_df$definition == "Primary solid Tumor", ])
normal_samples <- rownames(coldata_df[coldata_df$definition == "Solid Tissue Normal", ])

selected_cols <- c(sample(tumor_samples, 100), sample(normal_samples, 100))
mat_sub <- mat[, selected_cols]
df_sub <- data.frame(Type = coldata_df[selected_cols, "definition"])

```

# Plot the heatmap with the Viridis Palette
```{r}
library(ComplexHeatmap)
library(viridis)
library(circlize)

ha = HeatmapAnnotation(df = df_sub, 
                       col = list(Type = c("Primary solid Tumor" = "red2", 
                                           "Solid Tissue Normal" = "royalblue")))

# Viridis color mapping: Purple (Low), Teal (Mean), Yellow (High)
col_fun = colorRamp2(c(-2, 0, 2), viridis(3))

heatmap_final <- Heatmap(mat_sub, 
        name = "Z-score", 
        top_annotation = ha,
        cluster_rows = TRUE, 
        cluster_columns = TRUE,
        show_column_names = FALSE,
        column_km = 2,
        row_names_gp = gpar(fontsize = 5),
        column_title = "Molecular Signature: Top 50 BRCA Genes (n=200)",
        col = col_fun)

draw(heatmap_final, merge_legend = TRUE)

```


```{r}


```

